Inferring Lensing Mass in the Universe \\ from Photometric Catalog Data

### 2012-08-21 doc/acknowledgments.tex
first draft of theory section
notes on to-do items

TJC acknowledges support from STFC in the form of a research studentship.
%
PJM was given support by the TABASGO and Kavli foundations, and the Royal 
Society, in the form of research fellowships.
%
MWA \ldots. SH \ldots. SHS \ldots.
%
TT acknowledges support from the NSF through CAREER award NSF-0642621,
and from the Packard Foundation through a Packard Research Fellowship.
% 
LVEK acknowledges the support by an NWO-VIDI programme subsidy
(programme number 639.042.505).
%
This research is supported in part by the National Science Foundation under
Grant No. PHY99-07949.

>>>>

TEC acknowledges support from STFC in the form of a research studentship.
%
PJM was given support by the Kavli Foundation and the Royal 
Society, in the form of research fellowships.
%
MWA \ldots
%
SH \ldots
%
SHS \ldots
%
TT acknowledges support from the NSF through CAREER award NSF-0642621,
and from the Packard Foundation through a Packard Research Fellowship.
% 
LVEK acknowledges the support by an NWO-VIDI programme subsidy
(programme number 639.042.505).
%
This research is supported in part by the National Science Foundation under
Grant No. PHY99-07949.

### 2012-08-21 doc/paper-lcr.tex
notes on to-do items

Strong lenses sit in our real Universe -- they are not truely isolated;
there is some mass along the line of sight coming from intervening structures.
These structures may be physically associated with the lens e.g. other members
of a group or cluster, or they may be the result of chance alignments at any redshift. 
Whilst it is rare for three galaxies to line up well enough for both of the background sources
to be strongly lensed \citep{gavazzi2008,collett2012}, the shear scale of dark matter halos
makes a partial allignment -- with the additional structure acting as a weak lens -- 
a near certainty, indeed \citet{valewhite2003} and \citet{hilbert2007} find that there
are no empty lightcones in a realistic universe. These additional structures induce an external shear
which can be recovered in the mass-modelling of the lens, but they can also introduce a convergence that
is undetectable from the image positions, shapes and relative fluxes; {\it the mass sheet-degeneracy} \citep[see e.g.][for details]{Falco1985,Schneider2006}.

@@

@@

For real lines of sight, it is nearly impossible to know the true $\kappax$ for a
line of sight, and without a knowledge of $\kappax^{\mathrm{true}}$ it is hard to
quantify if there is a bias induced by using different $\kappax$ estimators.
A source with known unlensed brightness e.g. a strongly lensed type Ia
Supernova \flag{cite some SNe papers} or a source of known unlensed size \flag{ideas?}
would break the mass-sheet degeneracy and provide a value for $\kappax$, but
\flag{no?} such systems are currently known.

@@

With this catalogue of halos, positions, redshifts etc. it is possible to attempt
a reconstruction of the convergernce induced by every halo near a line of sight.
Cosmological dark matter simulations have shown that dark matter halos are well
approximated by NFW profiles \citep{NFW1997}, with a density profile given by
\be\label{eq:rhonfw}
\rho(r) = 
\frac{\rho_0}{\left(r/r_{s}\right)
\left(1+r/r_{s}\right)^{2}},
\ee
where $r_{s}$ is a characteristic radius of the cluster, representing where
the density slope transitions from $r^{-1}$ to $r^{-3}$. This radius is related
to the virial radius of the halo by $r_{s}~=~r_{200}/c$, where $r_{200}$ is the radius at 
which 
\be
\rho(r_{200}) = 200\rho_{{\mathrm{crit}}}(z), \text{ where } \rho_{\mathrm{crit}}(z) \equiv \frac{3H^{2}(z)}{8\pi G}
\ee
and $c$ is the concentration parameter, which can be estimated from the halo's mass,
using a mass--concentration relation. Typically more massive halos are less concentrated,
but there is some scatter; we use the relation of \citet{neto2007} to estimate $c$
from the mass enclosed within $r_{200}$, which we denote as $M_{200}$. The best fit relation of \citet{neto2007} is given by
\be
c_{200} = 4.67 (M_{200}/10^{14} h^{-1} \Msun)^{-0.11}.
\ee
$\rho_0$ is calculated from the critical density and concentration parameter:
\be
\rho_0 = \rho_{\mathrm{crit}}(z)\frac{200}{3}\frac{c^{3}}{\ln(1+c)-c/(1+c)}.
\ee

@@

@@

@@

@@

@@

Why reconstruction works or doesn't work.

Major sources of uncertainty.

Ideas for improving the procedure

Emphasize that this can be done before monitoring time is invested.

>>>>

Strong lenses sit in our real Universe -- they are not truely isolated;
there is some mass along the line of sight coming from intervening structures.
These structures may be physically associated with the lens e.g. other members
of a group or cluster, or they may be the result of chance alignments at any redshift. 
Whilst it is rare for three galaxies to line up well enough for both of the background sources
to be strongly lensed \citep{gavazzi2008,collett2012}, the shear scale of dark matter halos
makes a partial alignment -- with the additional structure acting as a weak lens -- 
a near certainty, indeed \citet{valewhite2003} and \citet{hilbert2007} find that there
are no empty lightcones in a realistic universe. These additional structures induce an external shear
which can be recovered in the mass-modelling of the lens, but they can also introduce a convergence that
is undetectable from the image positions, shapes and relative fluxes; {\it the mass sheet-degeneracy} \citep[see e.g.][for details]{Falco1985,Schneider2006}.

@@

\flag{The all-important calibration of the reconstruction has to be done
against Stefan's ray-traced kappas though -- so it is these that we have to
predict PDFs for, first and foremost. THEN we have to consider the question of
which kappa to use to correct the time delay distances: Stefan's kappa (as in
B1608 or RXJ1131), or kappa-keeton.}

@@

For real lines of sight, 
\phil{we would like to estimate the external convergence as best we can given all
data available. Typically this data is in the form of a photometric galaxy
catalog, containing positions, brightnesses, colours, and other type
information for each galaxy in the field of the lens, down to some magnitude
limit. Since we do not observe halo mass directy, we need some way estimating
the mass of each galaxy in the catalog, so that we can compute its
contribution to $\kappax$. This mass assignment recipe will be uncertain, but
can be calibrated to cosmological simulations.}

% PJM: What is this defeatism?!
% it is nearly impossible to know the true $\kappax$ for a
% line of sight, and without a knowledge of $\kappax^{\mathrm{true}}$ it is hard to
% quantify if there is a bias induced by using different $\kappax$ estimators.
%
% A source with known unlensed brightness e.g. a strongly lensed type Ia
% Supernova \flag{cite some SNe papers} or a source of known unlensed size \flag{ideas?}
% would break the mass-sheet degeneracy and provide a value for $\kappax$, but
% \flag{no?} such systems are currently known.

@@

\phil{Transforming the photometometric galaxy catalogue into a}  catalogue of
halo masses, positions, and redshifts, will enable us to attempt a
reconstruction of the convergence induced by every halo near a line of sight.
This is the subject of the first subsection below. We also need to account for
the mass in the Universe that is not associated with the galaxies we can see
-- this is the subject of the second subsection below.

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %

\subsection{Halos}

Cosmological dark matter simulations have shown that dark matter
halos are well approximated by NFW profiles \citep{NFW1997}, with a density
profile given by
\be\label{eq:rhonfw}
\rho(r) = 
\frac{\rho_0}{\left(r/r_{s}\right)
\left(1+r/r_{s}\right)^{2}},
\ee
where $r_{s}$ is a characteristic radius of the cluster, representing where
the density slope transitions from $r^{-1}$ to $r^{-3}$. This radius is related
to the virial radius of the halo by $r_{s}~=~r_{200}/c$, where $r_{200}$ is the radius at 
which 
\be
\rho(r_{200}) = 200\rho_{{\mathrm{crit}}}(z), \text{ where } \rho_{\mathrm{crit}}(z) \equiv \frac{3H^{2}(z)}{8\pi G}
\ee
and $c$ is the concentration parameter, which can be estimated from the halo's mass,
using a mass--concentration relation. Typically more massive halos are less concentrated,
but there is some scatter; we use the relation of \citet{neto2007} to estimate $c$
from the mass enclosed within $r_{200}$, which we denote as $M_{200}$. The best fit relation of \citet{neto2007} is given by
\be
c_{200} = 4.67 (M_{200}/10^{14} h^{-1} \Msun)^{-0.11}.
\ee
$\rho_0$ is calculated from the critical density and concentration parameter:
\be
\rho_0 = \rho_{\mathrm{crit}}(z)\frac{200}{3}\frac{c^{3}}{\ln(1+c)-c/(1+c)}.
\ee

@@

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %

\subsection{Voids}

@@

\flag{Need to draw a convergence from the void distribution, such that 1) the
intrinsic distribution of kappas is returned in the case where we have no
data, and 2) the mean density parameter over all lines of sight is $\Omega_m$.
The mass budget has to be balanced!}

@@

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %

\subsection{Comparison with Other Treatments of $\kappax$}
\label{sec:Dt:selection}

\phil{How does distance accuracy compare with simple averaging? eg assuming
$\kappax = 0$ for all lenses. How does it compare with the no data situation?
eg where the intrinsic distribution of kappa is used as $Pr(\kappax)$.}

@@

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %

\subsection{The Impact of Selection Effects}
\label{sec:Dt:selection}

\phil{How does distance accuracy vary with:}
\begin{itemize}
\item Depth of catalog? mlim = 22, 24, 26?
\item Overdensity cuts? Remove underdense fields, overdense fields (from
reconstructed kappa)
\item Shear-sensitive selection? What if we only selected quads?
\end{itemize}

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %

\subsection{Improving the Accuracy with More Information}
\label{sec:Dt:selection}

\phil{How does distance accuracy improve with:}
\begin{itemize}
\item Spectroscopic coverage? For galaxies above some mlim? For targets with
high kappa contribution (based on photometry)?
\item IR coverage for all objects? Improves photo-z and Mstar.
\item Including shear information from lens model? Assume accurate but
uncertain extertanl shear.
\end{itemize}

@@

Why reconstruction works or doesn't work.
Major sources of uncertainty.
Ideas for improving the procedure.
Emphasize that this can be done before monitoring time is invested.
